---
title: "WRBcalibrates Shiny App User Guide"
author: "Ghislain de Laplante"
date: "2023-05-31"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WRBcalibrates Shiny App User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# You should be modifying this vignette from the .Rmd document in the /vignettes folder, NOT the .RMD in the /doc folder.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(WRBcalibrates)

```

Important note regarding this vignette: if you are reading this in an HTML, Word or PDF document from the package folder on the G drive and wish to update the document, please do so via the vignette R Markdown document and re-publish/overwrite the document. The R Markdown document should be the ultimate source of truth regarding this R package and associated Shiny application. Instructions on how to do so are provided in the vignette Markdown document.

This vignette may also come to contain R code and links to other documentation. For best performance and to ensure that all links function, it is recommended to view this vignette from within R Studio. You can view all package vignettes with 'vignette(package = "WRBcalibrates")' or this particular vignette with 'vignette(topic = "WRBcalibrates_guide")'. Note however that vignettes may not have been installed, depending on the method you used to install the package.

```{r vignette building comments, eval=FALSE, include=FALSE}
# You should be modifying this vignette from the .Rmd document in the /vignettes folder, NOT the .RMD in the /doc folder.
# To have this vignette updated on the G drive, uncomment and run the following code *after* re-building the vignette using devtools::build_vignettes()
# file.copy(from = paste0(dirname(getwd()), "/doc/WRBcalibrates_guide.html"), to = "//env-fs/env-data/corp/water/Common_GW_SW/R-packages/WRBcalibrates/documentation/WRBcalibrates user guide.html", overwrite = TRUE)
```

# Introduction

The Water Resources Branch has a long and stellar history of deploying and using field instruments, but a less than excellent record-keeping methodology. Consequently, many historical field measurements dependent on accurate sensor readings are questionable at best, unusable at worst. To remedy this problem, new protocols call for regular calibrations or cross-checking of instruments where measurements are not corrected by field readings (e.g. water levels). Calibration records can then be used to a) validate field readings, and b) to apply corrections or offsets to field measurements, for example by shifting temperatures to account for linear sensor drift.

A first attempt at an application that could facilitate this record-keeping and, especially, facilitate future use of said data was done by CBTS in winter 2021. Unfortunately, due to poor software selection, staffing changes, and de-prioritization, this app was never fully operational. To avoid having another field season with only sparse paper records, a decision to create an in-house app was made in spring 2023, yielding an R Shiny application with much more flexibility and suitability for the task than the original Survey123 application.

The Shiny application is built within an R package, the intent of which is to facilitate the collection and storage instrument and calibration data, ensure safe storage, and facilitate future use of collected data. The calibrations app itself is deployed on the web at <https://YukonWRB/shinyapps.io/WRBcalibrates/> as of 2023-05-08 and interacts with Google sheets for data storage. The strengths and limitations of this system is discussed below, as are future improvements to app and data storage schemes.

# The WRBcalibrates package at a glance

The WRBcalibrates package is viewable at the private GitHub repository [YukonWRB/WRBcalibrates (github.com)](https://github.com/YukonWRB/WRBcalibrates). The repository should be viewable by members of the YukonWRB organization, and is built with a typical R package structure. The primary functions within this package are:

-   calConnect() establishes a connection to the PostgreSQL database where calibration data is stored. This will only work from within the YG network.

-   calibrate() starts the Shiny application

-   db_create() create an PostgreSQL database compliant with what is expected by the app.

The Shiny application itself lives in the isnt/app folder.

This package contains additional functions and capabilities, but these are beyond the scope of this vignette. Please consult the individual function help files or consult with the package developer for more information.

# Running and updating the Shiny app

You can run the app locally using function calibrate(), provided that you have YG network access and the necessary connection credentials in your .Renviron file. However, you will most likely always use the application from an internal or external-facing URL; see the Data Scientist for the URL.

Updating the Shiny app code locally is simple: install the latest version directly from GitHub using using function remotes::install_github(YukonWRB/WRBcalibrates) prior to calling function calibrate(). 


# Data storage

The application is set up to work with a PostgreSQL database with a structure identical to that created by function db_create(). This database back-end provides stability and security, and is more suitable for long-term storage than Google Sheets (which was the original setup).


# Future improvements

-   Adding a user-interface component that permits viewing calibration entries. It's a bit tricky to sort out how best to serve this information to the user, and with the Google Sheets back-end, how to do so quickly. Calibrations are spread out across 8 different tables: one for the basic observation information (observer, instrument, time, etc) and 7 for the individual parameters. An end-user might be looking for all calibrations on a certain day or all calibrations with a certain instrument (both of which entails searching each table) or might be looking for calibrations of a certain parameters (two tables; the basic observation, and the parameter table). Careful selection of the filters visible to the user is necessary.

-   The app could/should be deployed on a Linux machine using Shiny Server, as is done by the Yukon Geological Survey with their permafrost database. This solution bypasses would permit usage of the app without a direct or VPN connection to the YG network.
